A Formal Logic for Formal Category Theory

### 2022-10-14 paper/paper.tex
fix overflow
fubini examples done

\title{A Constructive Ordered Predicate Logic \\ for Category Theory}
\author{Max S. New\inst{1,2} and Daniel R. Licata\inst{2}}
\institute{University of Michigan \and Wesleyan University}

@@

Next, we have the ``Fubini'' theorems, which are analogous to simple
theorems relating tensor and hom types in ordered logic, and whose proofs are 
\begin{lemma}[{Fubini}]
  \label{lem:fubini-short}
  The following isomorphisms hold when the corresponding profunctors
  are well typed.
  \begin{enumerate}
  \item $\tensorexistsXwithYandZ {\beta} {\harrapp P \alpha \beta} {(\tensorexistsXwithYandZ \gamma {\harrapp Q \beta \gamma} {\harrapp R \gamma \delta})} \cong \tensorexistsXwithYandZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\harrapp R \gamma \delta}$
  \item $\homrallXYtoZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\harrapp S \alpha \gamma} \cong \homrallXYtoZ \beta {\harrapp P \alpha \beta} {\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {\harrapp S \alpha \gamma}}$
  \item $\homlallXYtoZ {\alpha} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\harrapp S \alpha \gamma} \cong \homlallXYtoZ \beta {\harrapp Q \beta \gamma} {\homlallXYtoZ \alpha {\harrapp P \alpha \beta} {\harrapp S \alpha \gamma}}$
  \item $\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {(\homlallXYtoZ \alpha {\harrapp P \alpha \beta} {\harrapp S \alpha \gamma})} \cong \homlallXYtoZ \alpha {\harrapp P \alpha \beta} {(\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {\harrapp S \alpha \gamma})}$
  \item $\pendallXdotY \alpha \homrallXYtoZ \beta P {P'} \cong \pendallXdotY \beta \homlallXYtoZ \alpha P {P'}$
  \end{enumerate}  
\end{lemma}
\begin{proof}
  Straightforward, for instance the forward direction of (1) is given
  by $\lambdaunary \alpha {x} \delta {\tensorelimWkontZ x
    {p.\beta.y. \tensorelimWkontZ y
      {q.\gamma.r. \tensorintroatXwithYandZ {\gamma}
        {\tensorintroatXwithYandZ \beta p q} {r}}}}$
\end{proof}

@@

Next, we provide some futher detailed examples of the formal category
theory constructions and theorems from Section~\ref{sec:examples}.

\begin{definition}[Synthetic Composition/Functoriality]
  We provide the definitions of the terms in \cref{construction:synthetic-composition}
  \begin{enumerate}
  \item Identity $\id = \pendlambdaXdotY \alpha \punitrefl \alpha$
  \item Composition $\textrm{comp} = \lambdaunary {\alpha_1} f {\alpha_2} \punitelimkontZatABC {\alpha. \homrallXYtoZ {\alpha_3} g g} {\alpha_1} f {\alpha_2}$
  \item Functoriality $\textrm{fctor}(F) = \lambdaunary {\alpha_1} f {\alpha_2} \punitelimkontZatABC {\alpha. \punitrefl{F\,\alpha}} {\alpha_1} f {\alpha_2}$
  \item Profunctoriality $\textrm{prof}(R) = \lambdaunary {\alpha_1} f {\alpha_2} {\punitelimkontZatABC {\alpha. \homrlambdaXatYdotZ r {\beta_1} \homrlambdaXatYdotZ g {\beta_2} {\homlappXtoYatZ {(\punitelimkontZatABC {\beta. \homllambdaXatYdotZ r \alpha r}{\beta_1}{g}{\beta_2})} {r} {\alpha}}}{\alpha_1} f {\alpha_2}}$

@@

@@

\section{Extended Synthetic Category Theory}

Next, we look at synthetic category theory constructions in SCT.

\subsection{Profunctor Isomorphisms}

We start our study of synthetic category theory constructions with
some isomorphisms that show the interaction between the 5
multiplicative connectives: right/left hom, end, unit and tensor.

First, we fix some terminology
\begin{lemma}
  Let $P, Q : \harr {\cat C} {\cat D}$.
  A homomorphism from $P$ to $Q$ is a term of type
  \[ \homunary {\alpha} {\harrapp P \alpha \beta} {\beta} {\harrapp Q \alpha \beta} \]
  Given homomorphisms $M : P \multimap Q$, $N : Q \multimap R$ we can define their composite as
  \[ (N \circ M) = \lambdaunary {\alpha} {x:\harrapp P \alpha \beta} \beta {\appunary \alpha N \beta {\appunary \alpha M \beta x}} \]
  This notion of composition is associative and unital using the
  $\beta\eta$ equations for the end/right hom connectives, with unit
  $\lambdaunary {\alpha} x \beta x$.

  An isomorphism between $P, Q : \harr {\cat C}{\cat D}$ is then a
  homomorphism $P \multimap Q$ with an inverse $Q \multimap P$.
\end{lemma}

Our first synthetic theorem is that we have that the unit, i.e., the
hom set, is the unit for tensor and hom.
%
The theorem for hom is the synthetic \emph{Yoneda lemma} and the
theorem for tensor is the synthetic \emph{Co-Yoneda lemma}.
\begin{lemma}[{Yoneda and Co-Yoneda}]
  \label{lem:yo-and-co-yo}
  Let $P : \harr {\cat C} {\cat D}$.

  For any $\gamma : \cat E \vdash b : \cat D$, the following are naturally isomorphic to $\harrapp P \alpha b$:
  \begin{enumerate}
  \item $\homrallXYtoZ {\beta} {(\punitinXfromYtoZ {D} {\beta} {b})} {\harrapp P {\alpha} {\beta}}$
  \item $\tensorexistsXwithYandZ {\beta} {\harrapp P \alpha {\beta}} {(\punitinXfromYtoZ {\cat D} {\beta} {b})}$
  \end{enumerate}

  For any $\gamma : \cat E \vdash a : \cat C$, the following are naturally isomorphic to $\harrapp P a \beta$:
  \begin{enumerate}
  \item $\homlallXYtoZ {{\alpha}} {(\punitinXfromYtoZ {C} {\alpha} {a})} {\harrapp P {{\alpha}} {\beta}}$
  \item $\tensorexistsXwithYandZ {\alpha} {(\punitinXfromYtoZ {C} {a} {\alpha})} {\harrapp P {\alpha}{\beta}}$
  \end{enumerate}
\end{lemma}
\begin{proof}
  TODO: what's better, build the substitution in or prove the
  substitution can be built in afterwards? The one building in the
  substitution obscures the connection to the normal tensor/hom/unit
  relationship, but the one with the built in subst is obviously more
  general.
\end{proof}

Next, we have various associativities between the homs, tensor and end
constructions. Semantically, these are related to so-called
\emph{Fubini} theorems about interchange of (co)-ends.
\begin{lemma}[{Fubini}]
  \label{lem:fubini}
  The following isomorphisms hold for any $P : \harr {\cat C}{\cat D},
  P' : \harr {\cat C}{\cat D} Q :\harr {\cat D}{\cat E}, R : \harr
  {\cat E} {\cat F}, S : \harr {\cat C} {\cat E}$. where $\alpha:\cat
  C, \beta:\cat D, \gamma : \cat E, \delta: \cat F$
  \begin{enumerate}
  \item $\tensorexistsXwithYandZ {\beta} {\harrapp P \alpha \beta} {(\tensorexistsXwithYandZ \gamma {\harrapp Q \beta \gamma} {\harrapp R \gamma \delta})} \cong \tensorexistsXwithYandZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\harrapp R \gamma \delta}$
  \item $\homrallXYtoZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\harrapp S \alpha \gamma} \cong \homrallXYtoZ \beta {\harrapp P \alpha \beta} {\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {\harrapp S \alpha \gamma}}$
  \item $\homlallXYtoZ {\alpha} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\harrapp S \alpha \gamma} \cong \homlallXYtoZ \beta {\harrapp Q \beta \gamma} {\homlallXYtoZ \alpha {\harrapp P \alpha \beta} {\harrapp S \alpha \gamma}}$
  \item $\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {(\homlallXYtoZ \alpha {\harrapp P \alpha \beta} {\harrapp S \alpha \gamma})} \cong \homlallXYtoZ \alpha {\harrapp P \alpha \beta} {(\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {\harrapp S \alpha \gamma})}$
  \item $\pendallXdotY \alpha \homrallXYtoZ \beta P {P'} \cong \pendallXdotY \beta \homlallXYtoZ \alpha P {P'}$
  \end{enumerate}  
\end{lemma}
\begin{proof}
  TODO
\end{proof}
\subsection{Composition, Functoriality, Profunctoriality, Naturality}

Hom sets don't provide primitive syntax for composition of morphisms
in a category. As our first synthetic construction, we can
\emph{define} composition using the elimination form for the unit.
%
\begin{definition}[Composition of Morphisms]
  We define composition of morphisms as the following term of type $\Pi_{\cat C:\Cat}\hombinary {\alpha_1} {\punitinXfromYtoZ {\cat C}{\alpha_1} {\alpha_2}} {\alpha_2} {\punitinXfromYtoZ {\cat C}{\alpha_2} {\alpha_3}} {\alpha_3} {\punitinXfromYtoZ {\cat C}{\alpha_1} {\alpha_3}}$:
  \[
  \lambda \cat C. \punitelimtoYwithkontZ {\hlambda {\alpha_1}{\alpha_2} \homrallXYtoZ {\alpha_3} {\punitinXfromYtoZ {\cat C} {\alpha_2} {\alpha_3}} {\punitinXfromYtoZ {\cat C} {\alpha_1} {\alpha_3}}} {\lambdaunary {\alpha} {f : \punitinXfromYtoZ {\cat C}{\alpha} {\alpha_3}} {\alpha_3} f}
  \]
  We call this term $\textrm{comp}$
\end{definition}
This has the same flavor as a directed type theory: to construct the
composite it is sufficient to define the action on the identity
morphism, which is of course the identity for composition.
%
Note that we gave a slight bias to the definition to follow by
induction on the \emph{first} function, whereas we could make the
following alternative definition, naturally of type
$\pendallXdotY {\alpha_2} \homrallXYtoZ {\alpha_3} {\punitinXfromYtoZ {\cat C}{\alpha_2} {\alpha_3}}
{(\homlallXYtoZ {{\alpha_1}} {\punitinXfromYtoZ {\cat C}{\alpha_1} {\alpha_2}}{\punitinXfromYtoZ {\cat C}{\alpha_1} {\alpha_3}})}$:
\[
  \punitelimtoYwithkontZ {\hlambda {\alpha_2}{\alpha_3} \homlallXYtoZ {{\alpha_1}} {\punitinXfromYtoZ {\cat C}{\alpha_1} {\alpha_2}}{\punitinXfromYtoZ {\cat C}{\alpha_1} {\alpha_3}} } {\lambdaunary {\alpha} {g : \punitinXfromYtoZ {\cat C}{\alpha_1}{\alpha}} {\alpha} g}
\]

\begin{lemma}[Equivalence of Composition Definitions]
  TODO
\end{lemma}

It should follow by $\beta\eta$ that this is associative and unital.

Next, let's prove synthetically that every functor $F : \varr{\cat
  C}{\cat D}$ is functorial in this notion of composition.
\begin{definition}
  For any $F : \varr {\cat C} {\cat D}$, we can define a morphism $F_1
  : \homunary {\alpha_1} {\punitinXfromYtoZ {\cat C} {\alpha_1}
    {\alpha_2}} {\alpha_2} {\punitinXfromYtoZ {\cat D} {F \alpha_1}
    {F \alpha_2}}$ as
  \[ F_1 = \punitelimtoYwithkontZ {\hlambda {\alpha_1}{\alpha_2} \punitinXfromYtoZ {\cat D} {F \alpha_1}{F \alpha_2}}
  {\pendlambdaXdotY \alpha \punitrefl {\cat D} {F \alpha}}
  \]
  Note that
  \[ A[\id_\alpha/\alpha] = \id_{A} \]
  \[ A[f;g] = A[f];A[g] \]
\end{definition}
Note the similarity to the proof that all functions are congruences in
MLTT/HoTT.

And similarly, every profunctor $R : \harr {\cat C}{\cat D}$ is
\emph{pro-functorial}.
\begin{definition}
  For any $R : \varr {\cat C} {\cat D}$, we can define left and right
  actions of composition:
  \[ \homunary {\alpha_1} {\tensorexistsXwithYandZ {\alpha_2} {(\punitinXfromYtoZ {\cat C} {\alpha_1}{\alpha_2})} {\tensorexistsXwithYandZ {\alpha_3} {(\harrapp R {\alpha_2} {\alpha_3})}}{(\punitinXfromYtoZ {\cat C} {\alpha_3}{\alpha_4})}} {\alpha_4} {\harrapp R {\alpha_1}{\alpha_4}}
  \]
  Furthermore we can show that they are functorial in that the induced functions
  TODO
  are equal
\end{definition}

Next we can define natural transformations as specific kinds of terms
whose type is $\pendallXdotY {\alpha} \punitinXfromYtoZ {\cat D} {F
  \alpha}{G \alpha}$. In fact we can generalize this to any
endo-profunctor $P$:
\begin{definition}[Natural element]
  A natural element of an endo-profunctor $P : \harr {\cat C}{\cat C}$
  is a term of type $\pendallXdotY {\alpha} \harrapp P \alpha \alpha$.

  A natural transformation is the special case where $P =
  \punitinXfromYtoZ {\cat D} {F \alpha}{G \alpha}$ for functors $F, G
  : \varr {\cat C}{\cat D}$.

  This is natural in the sense that the incuded terms
  \[ \homunary {\alpha_1}{\punitinXfromYtoZ {} {\alpha_1}{\alpha_2}} {\alpha_2} {\harrapp P {\alpha_1}{\alpha_2}}  \]
  and
  \[ \pendallXdotY {\alpha_2} \homlallXYtoZ {\alpha_1} {\punitinXfromYtoZ {} {\alpha_1}{\alpha_2}} {\harrapp P {\alpha_1}{\alpha_2}}\]

  are equal under the associativity equivalence between these two
  types.
\end{definition}

\subsection{Yoneda Lemma and Representable Presheaves}

Given any category $\cat C$, we can define the positive and negative
Yoneda embeddings $Y^\pm : \cat C \to \pmPresheaf {\cat C}$ using the
hom profunctor. Note that both of these functors are covariant in
$\cat C$, so semantically, $\posPresheaf {\cat C}$ should be
interpreted as $(\Set^{\cat C})^o$.

\begin{definition}
  Define the negative Yoneda embedding $Y^-[\alpha]$
  \[ \alpha:\cat C \vdash Y^-[\alpha] = \lambda \alpha':\cat C. \punitinXfromYtoZ {\cat C}{\alpha'}{\alpha} : \negPresheaf {\cat C} \]

  The positive Yoneda embedding is defined similarly $Y^+[\alpha]$.
\end{definition}

The Yoneda lemma says that the set of natural transformations from
$Y^-[\alpha]$ to a presheaf $p$ is naturally isomorphic to the
elements of the presheaf itself: $\negPresheafApp \alpha p$. In
encoding this statement in our type theory, we find that we can use
two different interpretations of ``natural transformations''.

On the one hand, a natural transformation should be a morphism in the
presheaf category. This interpretation of the Yoneda lemma would mean
that we have a natural isomorphism of profunctors:
\[ \alpha:\cat C; \pi:\negPresheaf {\cat C} \vdash \punitinXfromYtoZ{\negPresheaf {\cat C}} {Y^-[\alpha]} {\pi} \cong \negPresheafApp \alpha \pi \]
Attempting to prove this we become stuck very quickly because we know
nothing about the structure of morphisms in the presheaf category.

On the other hand, a natural transformation of presheaves $\pi$ and
$\rho$ can be defined as a family of functions from $\negPresheafApp
\alpha \pi$ to $\negPresheafApp \alpha \rho$, or in other words as an
\emph{end}. Specializing to the Yoneda lemma, we interpret ``a natural
transformation from $Y^-[\alpha]$ to $\pi$'' as the following, which simplifies using $\beta$ reduction:
\[ \alpha:\cat C; \pi:\negPresheaf {\cat C} \vdash \homrallXYtoZ {\alpha':\cat C} {\negPresheafApp {\alpha'} {Y^-[\alpha]}}{\negPresheafApp {\alpha'} \pi} = \homrallXYtoZ {\alpha':\cat C} {(\punitinXfromYtoZ {\cat C}{\alpha'}{\alpha})}{(\negPresheafApp {\alpha'} \pi)} \]
And then the Yoneda lemma becomes:
\[ \alpha:\cat C; \pi:\negPresheaf {\cat C} \vdash \homrallXYtoZ {\alpha':\cat C} {(\punitinXfromYtoZ {\cat C}{\alpha'}{\alpha})}{(\negPresheafApp {\alpha'} \pi)} \cong \negPresheafApp \alpha \pi\]
Which is an instance of our earlier Lemma\,\ref{lem:yo-and-co-yo}
which follows quite easily.

Then we are left with the question of whether these two forms of the
Yoneda lemma are equivalent, or slightly more generally are these two definitions of natural transformation of presheaves equivalent:
\[ \rho:\negPresheaf{\cat C}; \pi:\negPresheaf {\cat C} \vdash \punitinXfromYtoZ{\negPresheaf {\cat C}} {\rho} {\pi} \]
and
\[ \rho:\negPresheaf{\cat C}; \pi:\negPresheaf {\cat C} \vdash \homrallXYtoZ {\alpha:\cat C} {\negPresheafApp \alpha \rho}{\negPresheafApp \alpha \pi} \]
While we can show that there is a morphism from the hom set to the
end, the inverse does not seem provable, and seems analogous to the
\emph{univalence} axiom in homotopy type theory, as it tells us that
the morphisms of the universe are given by a particular kind of
function in our logic.

\subsection{Adjoint Functors and Representable Profunctors}

Next, we demonstrate that the two common definitions of adjoint
functors are equivalent: isomorphism of hom-sets and units and
co-units.
%
The proof goes through most easily, and utilizes duality most
effectively, if we factorize the construction through an intermediate
notion of when a functor \emph{represents} a profunctor.

\begin{definition}[Functor Represents a Profunctor]
  A functor $G : \varr{\cat D}{\cat C}$ right-represents a profunctor $R :
  \harr {\cat C}{\cat D}$ when there is an isomorphism of profunctors:
  \[ \punitinXfromYtoZ {\cat C} {\alpha} {G \beta} \cong \harrapp R {\alpha}{\beta}\]

  We say a functor $F : \varr{\cat C}{\cat D}$ \emph{left-represents} a profunctor $R : \harr {\cat C}{\cat D}$ when there is an isomorphism of profunctors
  \[ \punitinXfromYtoZ {\cat D} {F \alpha} {\beta} \cong \harrapp R {\alpha}{\beta}\]
\end{definition}

This is a parameterized generalization of the notion of an object
representing a functor $R : C^{op} \to Set$.
These notions are precisely dual. $F$ left-represents a profunctor $R$
precisely if it right-represents it in the opposite SCT.

Note that trivially any functor $G$ right-represents
$\hlambda\alpha\beta \punitinXfromYtoZ {\cat D}{\alpha} {G \beta}$ and
dually, any functor $F$ left-represents
$\hlambda\alpha\beta\punitinXfromYtoZ {\cat C}{F\alpha}{\beta}$.

Then we can show that any right-representation isomorphism is
determined by a co-unit:
\begin{lemma}[(Co)-unit of a Representation]
  If $G: \varr {\cat D}{\cat C}$ right-represents $R : \harr {\cat
    C}{\cat D}$, then there is a natural element, the co-unit:
  \[ \pendallXdotY \beta \harrapp R {G\beta} \beta \]
  Such that the half of the isomorphism 
  \[ \punitinXfromYtoZ {\cat C} {\alpha} {G \beta} \multimap \harrapp R {\alpha}{\beta} \]
  is defined by composition with the natural element.

  Dually, any $F$ left-representing isomorphism with $R$ is determined
  by a ``unit'' of type $\pendallXdotY \alpha \harrapp R {\alpha}
  {F\alpha}$.
\end{lemma}
\begin{proof}
  A homomorphism $\punitinXfromYtoZ {\cat C} {\alpha} {G \beta} \multimap \harrapp R {\alpha}{\beta}$ is by definition a term of type
  \[ \pendallXdotY \alpha \homrallXYtoZ {\beta} {(\punitinXfromYtoZ {\cat C} \alpha {G \beta})} {\harrapp R \alpha \beta} \]
  By \cref{lem:fubini}, this type is isomorphic to 
  \[ \pendallXdotY \beta \homlallXYtoZ {\alpha} {(\punitinXfromYtoZ {\cat C} \alpha {G \beta})} {\harrapp R \alpha \beta} \]
  By \cref{lem:yo-and-co-yo}, this type is isomorphic to
  \[ \pendallXdotY \beta {{\harrapp R {(G \beta)} \beta}} \]

  Therefore this half of the isomorphism is determined by a natural
  element. The construction in the proof of \cref{lem:yo-and-co-yo} is
  indeed by composition (i.e., unit elimination).

  The $F$ case is precisely dual.
\end{proof}

Combining these two, we get the unit-counit definition for adjoint
functors

\begin{definition}[Adjunction by unit/co-unit]
  We say $G : \varr {\cat D}{\cat C}$ is right adjoint to $F : \varr
  {\cat C}{\cat D}$ if there is a unit
  \[ \eta : \pendallXdotY \alpha \punitinXfromYtoZ \alpha {G(F(\alpha))}\]
  and a co-unit
  \[ \epsilon : \pendallXdotY \beta \punitinXfromYtoZ {F(G(\beta))} \beta \]

  satisfying the triangle identities...
  \[ \id_{R[\beta]} = R[\epsilon_{\beta}] \circ \eta_{R[\beta]} \]
  \[ \id_{L[\alpha]} = \epsilon_{L[\alpha]} \circ L[\eta_{\alpha}] \]
\end{definition}

%% \begin{theorem}[Adjunctions]
%%   $G$ is right adjoint to $F$ if and only if there is a natural isomorphism
%%   $\punitinXfromYtoZ {\cat D} {(F \alpha)} \beta \cong \punitinXfromYtoZ {\cat D} \alpha {G \beta}$
%% \end{theorem}
%% \begin{proof}
%%   First, by \cref{lem:co-unit}, the forward direction of the
%%   isomorphism is equivalent to providing a term
%%   \[ \eta : \pendallXdotY \alpha \punitinXfromYtoZ \alpha {G(F(\alpha)) } \]

%%   Second, the backward direction of the isomorphism is equivalent to
%%   providing a term
%%   \[ \epsilon : \pendallXdotY \beta \punitinXfromYtoZ {F(G(\beta))} \beta \]

%%   And the triangle laws state exactly that the induced functions are
%%   isomorphisms.
%% \end{proof}

%% \subsection{Kan Extensions/Lifts}

%% As a further example of the synthetic approach to universal
%% properties, we consider right and left \emph{Kan extensions}, which
%% are generalizations of limits and colimits, respectively.

%% The right Kan extension of $F$ along $J$ is the ``best'' (maximal) way to
%% fill in the following square:
%% \[% https://tikzcd.yichuanshen.de/#N4Igdg9gJgpgziAXAbVABwnAlgFyxMJZABgBpiBdUkANwEMAbAVxiRAGEQBfU9TXfIRQBGclVqMWbACLdeIDNjwEiZYePrNWiDgHJu4mFADm8IqABmAJwgBbJGRA4ISAEzVNUnQCk5lm-aIjs5IohJabABiBlxAA
%% \begin{tikzcd}
%% \cat C \arrow[d, "J"] \arrow[r, "F"] & \cat D \\
%% \cat C'                              &  
%% \end{tikzcd}\]
%% I.e. a functor $\textrm{Ran}_J F : \cat C' \to \cat D$ with a
%% universal natural transformation $\textrm{Ran}_J F \circ J \Rightarrow
%% F$. A left kan extension $\textrm{Lan}_J F$ is the same but with a
%% natural transformation $F \Rightarrow \textrm{Lan}_J F \circ J $.

%% We can express the concept in SCT as representing a certain profunctor
%% constructed using the profunctor hom:
%% \begin{definition}[Kan Extensions]
%%   Given functors $F : \varr {\cat C}{\cat D}$ and $J : \varr {\cat
%%     C}{\cat C'}$, a right Kan extension is a functor $\textrm{Ran}_J F
%%   : \varr {\cat C'}{\cat D} $that right-represents the profunctor
%%   \[ \hlambda {\beta:\cat D}{\alpha': \cat C} \homrallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{\alpha'}{J\alpha})} {\punitinXfromYtoZ {\cat D}{\beta}{F \alpha}} \]

%%   A left Kan extensions is a functor $\textrm{Lan}_J F : \varr {\cat
%%     C'}{\cat D}$ is a functor satisfying the dual property: it
%%   \emph{left}-represents the profunctor
%%   \[\hlambda {\alpha':\cat C}{\beta:\cat D} \homlallXYtoZ {\alpha}{(\punitinXfromYtoZ {\cat C'}{J\alpha}{\alpha'})} {\punitinXfromYtoZ {\cat D}{F \alpha}{\beta}}
%%   \]
%% \end{definition}

>>>>

\title{A Formal Logic for Formal Category Theory}
\author{Max S. New\inst{1,2} and Daniel R. Licata\inst{2}}
\institute{University of Michigan \and Wesleyan University}

@@

Next, we have the ``Fubini'' theorems, which are analogous to simple
theorems relating tensor and hom types in ordered logic, and whose proofs are 
\begin{lemma}[{Fubini}]
  \label{lem:fubini-short}
  The following isomorphisms hold when the corresponding profunctors
  are well typed.
  \begin{enumerate}
  \item $\tensorexistsXwithYandZ {\beta} {\harrapp P \alpha \beta} {(\tensorexistsXwithYandZ \gamma {\harrapp Q \beta \gamma} {\harrapp R \gamma \delta})} \cong \tensorexistsXwithYandZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \alpha \beta} {\harrapp Q \beta \gamma})} {\harrapp R \gamma \delta}$
  \item $\homrallXYtoZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \delta \beta} {\harrapp Q \beta \gamma})} {\harrapp S \alpha \gamma} \cong \homrallXYtoZ \beta {\harrapp P \delta \beta} {\homrallXYtoZ \gamma {\harrapp Q \beta \gamma} {\harrapp S \alpha \gamma}}$
  \item $\homlallXYtoZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \gamma \beta} {\harrapp Q \beta \alpha})} {\harrapp S \gamma \delta} \cong \homlallXYtoZ \beta {\harrapp Q \beta \alpha} {\homlallXYtoZ \gamma {\harrapp P \gamma \beta} {\harrapp S \gamma \delta}}$
  \item $\homrallXYtoZ \gamma {\harrapp Q \delta \gamma} {(\homlallXYtoZ \beta {\harrapp P \beta \alpha} {\harrapp S \beta \gamma})} \cong \homlallXYtoZ \beta {\harrapp P \beta \alpha} {(\homrallXYtoZ \gamma {\harrapp Q \delta \gamma} {\harrapp S \beta \gamma})}$
  \item $\pendallXdotY \alpha \homrallXYtoZ \beta {\harrapp P \alpha \beta} {\harrapp Q \alpha \beta} \cong \pendallXdotY \beta \homlallXYtoZ \alpha {\harrapp P \alpha \beta} {\harrapp Q \alpha \beta}$
  \end{enumerate}  
\end{lemma}
\begin{proof}
  Straightforward, for instance the forward direction of (1) is given
  by $\lambdaunary \alpha {x} \delta {\tensorelimWkontZ x
    {p.\beta.y. \tensorelimWkontZ y
      {q.\gamma.r. \tensorintroatXwithYandZ {\gamma}
        {\tensorintroatXwithYandZ \beta p q} {r}}}}$
\end{proof}

@@

Next, we provide some futher details for some of the examples of the
formal category theory constructions and theorems from
Section~\ref{sec:examples}.

\begin{definition}[Synthetic Composition/Functoriality]
  We provide the definitions of the terms in Construction~\ref{construction:synthetic-composition}
  \begin{enumerate}
  \item Identity $\id = \pendlambdaXdotY \alpha \punitrefl \alpha$
  \item Composition $\textrm{comp} = \lambdaunary {\alpha_1} f {\alpha_2} \punitelimkontZatABC {\alpha. \homrallXYtoZ {\alpha_3} g g} {\alpha_1} f {\alpha_2}$
  \item Functoriality $\textrm{fctor}(F) = \lambdaunary {\alpha_1} f {\alpha_2} \punitelimkontZatABC {\alpha. \punitrefl{F\,\alpha}} {\alpha_1} f {\alpha_2}$
  \item Profunctoriality
    \[\textrm{prof}(R) = \lambdaunary {\alpha_1} f {\alpha_2} {\punitelimkontZatABC {\alpha. \homrlambdaXatYdotZ r {\beta_1} \homrlambdaXatYdotZ g {\beta_2} {\homlappXtoYatZ {(\punitelimkontZatABC {\beta. \homllambdaXatYdotZ r \alpha r}{\beta_1}{g}{\beta_2})} {r} {\alpha}}}{\alpha_1} f {\alpha_2}}\]

@@

\begin{lemma}[Fubini]
  We show two of the Fubini cases in detail:

  \begin{enumerate}
  \item $\homlallXYtoZ {\gamma} {(\tensorexistsXwithYandZ \beta {\harrapp P \gamma \beta} {\harrapp Q \beta \alpha})} {\harrapp S \gamma \delta} \cong \homlallXYtoZ \beta {\harrapp Q \beta \alpha} {\homlallXYtoZ \gamma {\harrapp P \gamma \beta} {\harrapp S \gamma \delta}}$
  \item $\pendallXdotY \alpha \homrallXYtoZ \beta {\harrapp P \alpha \beta} {\harrapp Q \alpha \beta} \cong \pendallXdotY \beta \homlallXYtoZ \alpha {\harrapp P \alpha \beta} {\harrapp Q \alpha \beta}$
  \end{enumerate}
\end{lemma}
\begin{proof}
  \begin{enumerate}
  \item This is a form of Currying isomorphism, as the $\lambda$ term
    makes clear:
    \begin{itemize}
    \item Left to Right
      \[ \lambdaunary \alpha h \delta \homllambdaXatYdotZ q \beta \homllambdaXatYdotZ p \gamma \homlappXtoYatZ h {(p,\beta,q)} \gamma  \]
    \item Right to Left
      \[ \lambdaunary {\alpha} k \delta \homllambdaXatYdotZ w \gamma \tensorelimWkontZ {w} {p.\beta.q. \homlappXtoYatZ {\homlappXtoYatZ k p \gamma} q \beta } \]
    \item Left to Right to Left
      \begin{align*}
        &\lambdaunary \alpha h \delta \homllambdaXatYdotZ w \gamma {\tensorelimWkontZ {w} {p.\beta.q. (\homlappXtoYatZ {\homlappXtoYatZ {(\homllambdaXatYdotZ q \beta \homllambdaXatYdotZ p \gamma {\homlappXtoYatZ h {(p,\beta,q)} \gamma})} q \beta }  p \gamma) }}\\
        &=\lambdaunary \alpha h \delta \homllambdaXatYdotZ w \gamma {\tensorelimWkontZ {w} {p.\beta.q. (\homlappXtoYatZ {{(\homllambdaXatYdotZ p \gamma {\homlappXtoYatZ h {(p,\beta,q)} \gamma})} }  p \gamma) }}\tag{contrahom$\beta$}\\
        &=\lambdaunary \alpha h \delta \homllambdaXatYdotZ w \gamma {\tensorelimWkontZ {w} {p.\beta.q. {{({\homlappXtoYatZ h {(p,\beta,q)} \gamma})} } }}\tag{contrahom$\beta$}\\
        &=\lambdaunary \alpha h \delta \homllambdaXatYdotZ w \gamma \homlappXtoYatZ h w \gamma\tag{tensor$\eta$}\\
        &=\lambdaunary \alpha h h\tag{contrahom$\eta$}
      \end{align*}
    \item Right to Left to Right
      \begin{align*}
        &\lambdaunary \alpha k \delta \homllambdaXatYdotZ q \beta \homllambdaXatYdotZ p \gamma \homlappXtoYatZ {(\homllambdaXatYdotZ w \gamma {\tensorelimWkontZ {w} {p.\beta.q. \homlappXtoYatZ {\homlappXtoYatZ k q \beta}  p \gamma }})} {(p,\beta,q)} \gamma\\
        &=\lambdaunary \alpha k \delta \homllambdaXatYdotZ q \beta \homllambdaXatYdotZ p \gamma { {\tensorelimWkontZ { {(p,\beta,q)}} {p.\beta.q. \homlappXtoYatZ {\homlappXtoYatZ k q \beta } p \gamma }}}\\
        &=\lambdaunary \alpha k \delta \homllambdaXatYdotZ q \beta \homllambdaXatYdotZ p \gamma {\homlappXtoYatZ {\homlappXtoYatZ k q \beta } p \gamma}\\
        &=\lambdaunary \alpha k \delta \homllambdaXatYdotZ q \beta { {\homlappXtoYatZ k q \beta}}\\
        &=\lambdaunary \alpha k \delta k\\
      \end{align*}
    \end{itemize}
  \item This isomorphism relates left and right homs. Unlike the
    previous cases, the isomorphism is of types, not sets/profunctors.
    \begin{itemize}
    \item Left to right
      \[\lambda X. \pendlambdaXdotY \beta \homllambdaXatYdotZ p \alpha {\appunary X \alpha p \beta}\]
    \item Right to left
      \[ \lambda Y. \pendlambdaXdotY \alpha \homrlambdaXatYdotZ p \beta {\homlappXtoYatZ {\pendappXtoY Y \beta} p \alpha} \]
    \item Left to right to left
      \begin{align*}
        &\lambda X. \pendlambdaXdotY \alpha \homrlambdaXatYdotZ p \beta {\homlappXtoYatZ {\pendappXtoY {(\pendlambdaXdotY \beta \homllambdaXatYdotZ p \alpha {\appunary X \alpha p \beta})} \beta} p \alpha}\\
        &= \lambda X. \pendlambdaXdotY \alpha \homrlambdaXatYdotZ p \beta {\homlappXtoYatZ { {(\homllambdaXatYdotZ p \alpha {\appunary X \alpha p \beta})}} p \alpha}\tag{nat.elt.$\beta$}\\
        &= \lambda X. \pendlambdaXdotY \alpha \homrlambdaXatYdotZ p \beta {\appunary X \alpha p \beta}\tag{contrahom$\beta$}\\
        &= \lambda X. \pendlambdaXdotY \alpha {\pendappXtoY X \alpha}\tag{contrahom$\eta$}\\
        &= \lambda X. X\tag{nat.elt.$\eta$}\\
      \end{align*}
    \item The other case is similar.
    \end{itemize}
  \end{enumerate}
\end{proof}

@@
